---
title: "Predicting Total Residential Building Permits in British Columbia"
date: "2024-12-02"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(fpp3)
library(readxl) # For reading Excel files
library(tidyr) 
library(purrr)
library(quantmod)
library(fable.prophet)
library(prophet)
library(kableExtra)
library(webshot2)
```

```{r}
# Step 1: Read the data
data <- read_excel("DATA/BUILDING_PERMITS_CANADA.xlsx", sheet = 1)

# View the first few rows of the dataset
head(data)

# Reshape the data from wide to long format
tidy_data <- data %>%
  pivot_longer(
    cols = starts_with("Jan-"), # Adjust to match the actual column patterns
    names_to = "DATE",
    values_to = "Value"
  )

# Separate the columns into specific data types if necessary
tidy_data <- tidy_data %>%
  rename(
    `Types of work, total` = `Type of work 6`,
    `Seasonal adjustment, value type` = `Seasonal adjustment, value type 4 5`,
    `Geography` = Geography
  )

# Finalize the format
tidy_data <- tidy_data %>%
  select(
    `Types of work, total`,
    `Seasonal adjustment, value type`,
    Geography,
    DATE,
    everything()
  )

# View the reshaped data
print(tidy_data)

# Step 2: Identify date columns (adjust based on your data)
# Assuming all date columns start with a month abbreviation (e.g., "Jan-", "Feb-", etc.)
date_columns <- grep("^(Jan|Feb|Mar|Apr|May|Jun|Jul|Aug|Sep|Oct|Nov|Dec)-", names(data), value = TRUE)

# Step 3: Reshape data from wide to long format
tidy_data <- data %>%
  pivot_longer(
    cols = all_of(date_columns),   # Only reshape the date columns
    names_to = "DATE",             # New column for dates
    values_to = "Value"            # New column for values
  )

# Step 4: Clean and reformat DATE column
# Assuming dates are formatted as "MMM-YY" (e.g., "Jan-23"), convert them to proper date format
tidy_data <- tidy_data %>%
  mutate(
    Value = as.numeric(Value),
    DATE = as.Date(paste0("01-", DATE), format = "%d-%b-%y")  # Adding "01-" assumes day 1 for month-year
  )

# Step 5: Reorder columns to match the desired output
tidy_data <- tidy_data %>%
  select(
    #`Type of work 6`,
    #`Seasonal adjustment, value type 4 5`,
    #File,
    Geography,
    DATE,
    Value,
    Variables,
    `Type of building`
  )

# Step 6: Save the transformed data to a new CSV file
write.csv(tidy_data, "Transformed_Building_Permits.csv", row.names = FALSE)

# Preview of table properties
print(glimpse(tidy_data))

# Preview the transformed data
print(knitr::kable(head(tidy_data)))

print(knitr::kable(
  tidy_data |>
    filter(`Type of building` %in% c('Total non-residential','Total residential','Total residential and non-residential')) |>
    group_by(Geography,`Type of building`,Variables) |>
    summarise(Value = sum(Value), Count = n()) |>
    mutate(Value = format(Value,big.mark = ',')) |>
    select(Geography,`Type of building`,Variables, Value, Count)
))
```


```{r}
# Choose just the data we need to plot.
filtered_data <- tidy_data |>
  filter(`Type of building` %in% c('Total non-residential', 'Total residential', 'Total residential and non-residential')) |> 
  filter(`Type of building` %in% c('Total residential')) |> 
  #filter(`Geography` %in% c('Alberta','British Columbia','Ontario')) |> 
  filter(`Variables` %in% c('Number of permits')) |> 
  filter(`Geography` %in% c('British Columbia')) |> 
  mutate(DATE = yearmonth(DATE)) |>
  select(Geography,`Type of building`,Variables, Value, DATE) |>
  #arrange(`Type of building`,Geography,Variables) |>
  as_tsibble(index = DATE, key = c(Geography,`Type of building`,Variables))

head(filtered_data)

# Loop to generate and display plots for each category
for (var in unique(filtered_data$Variables)) {
  for (type in unique(filtered_data$`Type of building`)) {
    for (geo in unique(filtered_data$Geography)) {
      # Subset the data for the current combination
      subset_data <- filtered_data |>
        filter(Geography == geo, `Type of building` == type, Variables == var)

      # Generate the plot
      
      p <- autoplot(subset_data, Value) +
        geom_line(color = "blue") + 
        labs(title = paste("Geography:", geo),y = var, x = type) +
        theme_minimal()
      
      #par(mfrow = c(1, 2))
      # Print the plot
      print(p)
      # par(mfrow = c(1,1))
      
      ggsave(paste0("images/","01_", geo, "_", type, "_", var, "_plot.jpg"), plot = p, width = 8, height = 6,create.dir = TRUE)
      
      data <- subset_data

      #2. Split Data
      #Split the data into training and testing sets for evaluation:
      train_data <- data %>% filter(DATE < yearmonth("2024 May"))
      test_data <- data %>% filter(DATE >= yearmonth("2024 May"))
    }
  }
}
```

## ARIMA
### Check for Stationarity
```{r}
# Plot ACF and PACF
acf_plot <- ACF(train_data, Value) %>% autoplot() +
  labs(title = "ACF Train Data") +
  theme_minimal()

pacf_plot <- PACF(train_data, Value) %>% autoplot() +
  labs(title = "PACF Train Data") +
  theme_minimal()

ggsave(paste0("images/","02_acf_plot.jpg"), plot = acf_plot,create.dir = TRUE)
ggsave(paste0("images/","02_pacf_plot.jpg"), plot = pacf_plot,create.dir = TRUE)

print(acf_plot)
print(pacf_plot)

# Perform stationarity test
library(urca)
adf_test <- ur.df(train_data$Value, type = "drift")
summary(adf_test)
```

### Fit ARIMA and Benchmark Models
```{r}
# Fit ARIMA model
arima_model <- train_data %>% model(ARIMA(Value))

# Fit benchmark models
benchmark_models <- train_data %>% model(
  Mean = MEAN(Value),
  Naive = NAIVE(Value),
  Drift = RW(Value ~ drift())
)

mean_Model <- train_data %>% model(
  Mean = MEAN(Value)
)

naive_Model <- train_data %>% model(
  Naive = NAIVE(Value)
)

drift_Model <- train_data %>% model(
  Drift = RW(Value ~ drift())
)
```

### Evaluate Models
```{r}
# Forecast using all models
forecasts <- bind_rows(
  arima_model %>% forecast(h = nrow(test_data)),
  benchmark_models %>% forecast(h = nrow(test_data))
)

# Evaluate accuracy
arima_accuracy_metrics <- forecasts %>%
  accuracy(test_data)

# Display evaluation
print(arima_accuracy_metrics)
```

### Visualize forecasts and actual values
```{r}
arima_forecast_plot <-
forecasts %>%
  autoplot(data) +
  labs(title = "Forecast vs Actual",
       y = "Permits", x = "Date") +
  theme_minimal()

ggsave(paste0("images/","03_arima_forecast_plot.jpg"), plot = arima_forecast_plot,create.dir = TRUE)

print(arima_forecast_plot)

```


### Residual Diagnostics for ARIMA and Benchmark Models
```{r}
plot <- gg_tsresiduals(drift_Model)
ggsave("images/04_drift_residuals_plot.png", plot, width = 8, height = 6)
print(plot)
# png("images/04_drift_residuals_plot.png", width = 800, height = 600)
# gg_tsresiduals(drift_Model)
# dev.off()
```

### Statistical Tests: Box-Pierce and Ljung-Box
```{r}
# Augment model to extract residuals
drift_aug <- drift_Model |> augment()

# Perform Box-Pierce Test
box_pierce_results <- drift_aug |>
  features(.innov, box_pierce, lag = 10) |>
  rename(Box_Pierce_pvalue = bp_pvalue)

# Perform Ljung-Box Test
ljung_box_results <- drift_aug |>
  features(.innov, ljung_box, lag = 10) |>
  rename(Ljung_Box_pvalue = lb_pvalue)

# Combine the Results
residual_tests <- bind_cols(
  box_pierce_results |> select(Box_Pierce_pvalue),
  ljung_box_results |> select(Ljung_Box_pvalue)
)

# Display the Results
print(knitr::kable(residual_tests, caption = "Residual Diagnostic Tests (Box-Pierce & Ljung-Box)"))
```

## EXPONENTIAL SMOOTHING
### Fit Models
```{r}
#Fit various exponential smoothing models to the training data.
#Simple Exponential Smoothing (SES):
ses_model <- train_data %>%
model(SES = ETS(Value ~ error("A") + trend("N") + season("N")))

#Holt’s Linear Trend Method:
holt_model <- train_data %>%
model(Holt = ETS(Value ~ error("A") + trend("A") + season("N")))

#Holt-Winters Seasonal Methods:
hw_additive_model <- train_data %>%
model(HoltWinters_Additive = ETS(Value ~ error("A") + trend("A") + season("A")))
hw_multiplicative_model <- train_data %>%
model(HoltWinters_Multiplicative = ETS(Value ~ error("M") + trend("A") + season("M")))
```

### Forecast Future Values
```{r}
#Generate forecasts for the test period:
forecasts <- bind_rows(
  ses_model %>% forecast(h = nrow(test_data)),
  holt_model %>% forecast(h = nrow(test_data)),
  hw_additive_model %>% forecast(h = nrow(test_data)),
  hw_multiplicative_model %>% forecast(h = nrow(test_data))
)
```

### Evaluate Models
```{r}
#Calculate forecast accuracy using metrics like RMSE, MAE, and MAPE:
exponential_smoothing_accuracy_metrics <- forecasts %>%
  accuracy(test_data)

print(exponential_smoothing_accuracy_metrics)
```

### Visualize Results
```{r}
#Plot the actual values and forecasts for each method:
exponential_smooth_forecast_plot <- forecasts %>%
autoplot(data) +
labs(title = "Exponential Smoothing Forecast Comparisons", y = "Number of Permits", x = "Date") +
theme_minimal()

ggsave(paste0("images/","05_exponential_smooth_forecast_plot.jpg"), plot = exponential_smooth_forecast_plot,create.dir = TRUE)

print(exponential_smooth_forecast_plot)
#print(forecasts)
```

## NEURAL NETWORK AUTOREGRESSION (NNETAR) AND PROPHET
### NNETAR
```{r}
nnetar_model <- train_data %>%
  model(NNETAR = NNETAR(sqrt(Value)))

# Forecast with NNETAR Model
nnetar_forecast <- nnetar_model %>%
  forecast(h = nrow(test_data))

# Plot NNETAR Forecast
nnetar_forecast_plot <- autoplot(nnetar_forecast, data) +
  labs(title = "NNETAR Forecast", y = "Building Permits", x = "Date") +
  theme_minimal()

ggsave(paste0("images/","06_nnetar_forecast_plot.jpg"), plot = nnetar_forecast_plot,create.dir = TRUE)

print(nnetar_forecast_plot)

nnetar_accuracy <- nnetar_forecast %>%
  accuracy(test_data)

# Combine Accuracy Metrics
nnetar_accuracy <- bind_rows(
  mutate(nnetar_accuracy, .model = "NNETAR")
)

# Display Accuracy Metrics
print(knitr::kable(nnetar_accuracy, caption = "NNETAR Model Accuracy"))
```

### Prophet model
```{r}
# Define the training dataset with proper dates
N <- nrow(data)
n <- 5 # Number of periods to forecast

# Create the training data
train_dates <- data$DATE[1:(N - n)] # Use actual dates
train_values <- data$Value[1:(N - n)]

# Create a data frame for Prophet
df_prophet <- data.frame(ds = train_dates, y = train_values) # ds = dates, y = values
head(df_prophet) # Verify the structure

# Fit the Prophet model
m <- prophet(df_prophet)

# Generate future dates starting after the last date in the training data
future_prophet <- make_future_dataframe(m, periods = n, freq = "month")
tail(future_prophet) # Verify the future dates

# Forecast the future values
forecast_prophet <- predict(m, future_prophet)

forecast_data <- forecast_prophet %>%
  inner_join(test_data, by = c("ds" = "DATE")) %>%  # Match forecast dates with test_data dates
  select(ds, yhat, yhat_lower, yhat_upper)  # Select necessary columns from the forecast

# Create the plot
prophet_forecast_plot <- ggplot() +
  # Plot training data
  geom_line(data = df_prophet, aes(x = ds, y = y, color = "Training Data"), size = 1, alpha = 0.6) +
  # Plot actual test data
  geom_line(data = test_data, aes(x = DATE, y = Value, color = "Actual Test Data"), size = 1, alpha = 0.8) +
  # Plot forecasted values
  geom_line(data = forecast_data, aes(x = ds, y = yhat, color = "Forecasted Values"), size = 1) +
  # Add confidence interval
  geom_ribbon(data = forecast_data, aes(x = ds, ymin = yhat_lower, ymax = yhat_upper, fill = "Confidence Interval"), alpha = 0.3) +
  # Customize colors and labels for the legend
  scale_color_manual(
    name = "Legend",
    values = c("Training Data" = "black", "Actual Test Data" = "blue", "Forecasted Values" = "red")
  ) +
  scale_fill_manual(
    name = "Legend",
    values = c("Confidence Interval" = "grey")
  ) +
  # Add title and labels
  ggtitle("Prophet Forecast with Actual and Training Data") +
  labs(
    x = "Date",
    y = "Building Permits"
  ) +
  theme_minimal() +
  theme(legend.position = "right")  # Place legend at the right

ggsave("images/07_forecast_prophet_plot.jpg", plot = prophet_forecast_plot, width = 8, height = 6)

print(prophet_forecast_plot)

#Plot forecast components (trend, seasonality)
print(prophet_plot_components(m, forecast_prophet))  # Generate the base plot

# Plot forecast components (trend, seasonality)
png("images/07_prophet_components_plot.png", width = 800, height = 600)
prophet_plot_components(m, forecast_prophet)
dev.off()

# Extract the test data
test_dates <- test_data$DATE # Dates of the test data
test_values <- test_data$Value # Actual values of the test data

# Subset the forecast to match the test dates
forecast_subset <- forecast_prophet %>%
  filter(as.Date(ds) %in% as.Date(test_dates)) %>%
  select(ds, yhat)

# Combine actual and forecasted values
performance_data <- data.frame(
  ds = test_dates,
  actual = test_values,
  predicted = forecast_subset$yhat
)

# Calculate accuracy metrics
rmse <- sqrt(mean((performance_data$actual - performance_data$predicted)^2))
mae <- mean(abs(performance_data$actual - performance_data$predicted))
mape <- mean(abs((performance_data$actual - performance_data$predicted) / performance_data$actual)) * 100

# Display the results
prophet_accuracy <- data.frame(
  .model = "PROPHET",
  RMSE = rmse,
  MAE = mae,
  MAPE = mape
)

print(knitr::kable(prophet_accuracy, caption = "Prophet Model Accuracy"))
```


## ACCURACY COMPARISION MODELS PERFORMANCE
### Merge and print the accuracy metrics

```{r}
# Merge the accuracy metrics
joined_accuracy_metrics <- bind_rows(
  arima_accuracy_metrics,
  exponential_smoothing_accuracy_metrics,
  prophet_accuracy,
  nnetar_accuracy
)

# Select key metrics and arrange by RMSE
evaluation_summary <- joined_accuracy_metrics %>%
  select(.model, RMSE, MAE, MAPE) %>%
  arrange(RMSE)

# Display the summary in a table
png("images/08_model_evaluation_summary.png", width = 800, height = 600)

# Open a new plot window
grid::grid.newpage()

# Print the table using knitr::kable
grid::grid.draw(
  gridExtra::tableGrob(
    evaluation_summary,
    theme = gridExtra::ttheme_default(core = list(fg_params = list(cex = 0.8))),
    rows = NULL
  )
)

dev.off()

print(knitr::kable(evaluation_summary, caption = "Model Evaluation Summary"))
```
